Project Description -- CAREER: Erosion, Seepage, Transport, and Collapse in Granular and Porous Media

### 2019-11-10 docs/proposals/NSF/ProjDesc/ProjDesc.tex
starting to pick away at proposal

%---------------------------------------------------------------------------------------------------%
%% PACKAGES
\usepackage{fullpage}
%\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx, amsmath, amssymb, amsfonts, mathtools, mathrsfs, color}
\usepackage{comment, enumerate, tabularx}
\usepackage{natbib, hyperref, url}
%\usepackage[justification=RaggedRight]{caption}
\usepackage{outlines,enumitem}
%---------------------------------------------------------------------------------------------------%

%---------------------------------------------------------------------------------------------------%
%% LATEX DEFINITIONS
% Basic editing
\newcommand{\nick}[1]{{\color{red}#1}}
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\here}{{\color{blue} here}}
% Basic math, derivatives
\newcommand{\td}[2]{\frac{d #1 }{d #2}}
\newcommand{\ttd}[2]{\frac{d^2 #1 }{{d #2}^2}}
\newcommand{\pd}[2]{ \frac{ \partial #1}{ \partial #2 } }
\newcommand{\ppd}[2]{ \frac{ \partial^2 #1}{ {\partial #2}^2 } }
\newcommand{\mppd}[3]{ \partial_{#2 #3} #1 }
% Basic math, vectors and other
%\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\bvec}[1]{{\mathbf{#1}}}
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\tavg}[1]{\langle #1 \rangle}
\newcommand{\norm}[1]{ \left\| #1 \right\| }
\newcommand{\ip}[1]{ \langle #1 \rangle }
\newcommand{\eps}{\varepsilon}
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}

@@

The first problem discussed here is the erosion of a porous medium, as has been preliminarily investigated in \cite{Quaife2018}. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, erosion results from surface shear stress
\begin{align}
\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations \eqref{StokesEq}. 
The motion of the interface is proportional to the absolute value of shear
\begin{align}
\Vn = \CE \, \abs{\tau},
\end{align}
In the later case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
\label{Ceq}
& \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C \, ,
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \left. \Diff \pd{C}{\ny} \right. %\right\vert_{\bdry} \, ,
\end{equation}
Both processes have been investigated by the PI and collaborators in the context of single bodies in high Reynolds-number flows \cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the low-Reynolds regime of porous-media flow, numerical methods have been developed by the PI and collaborators to simulate shear-induced erosion \cite{Quaife2018}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques. The boundary integral solver is accelerated by the fast multipole method \cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes.
%A near-term goal is to adapt these methods to handle the case of chemically-induced dissolution.

@@

\subsection{Sinkhole modeling}

>>>>

%---------------------------------------------------------------------------------------------------%
%% PACKAGES
\usepackage{fullpage}
%\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx, amsmath, amssymb, amsfonts, mathtools, mathrsfs, color}
\usepackage{comment, enumerate, tabularx}
\usepackage{natbib, hyperref, url}
%\usepackage[justification=RaggedRight]{caption}
\usepackage{outlines,enumitem}
\usepackage{wrapfig}
%---------------------------------------------------------------------------------------------------%

%---------------------------------------------------------------------------------------------------%
%% LATEX DEFINITIONS
% Basic editing
\newcommand{\nick}[1]{{\color{red}#1}}
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\here}{{\color{blue} here}}
% Basic math, derivatives
\newcommand{\td}[2]{\frac{d #1 }{d #2}}
\newcommand{\ttd}[2]{\frac{d^2 #1 }{{d #2}^2}}
\newcommand{\pd}[2]{ \frac{ \partial #1}{ \partial #2 } }
\newcommand{\ppd}[2]{ \frac{ \partial^2 #1}{ {\partial #2}^2 } }
\newcommand{\mppd}[3]{ \partial_{#2 #3} #1 }
% Basic math, vectors and other
%\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\bvec}[1]{{\mathbf{#1}}}
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\tavg}[1]{\langle #1 \rangle}
\newcommand{\norm}[1]{ \left\| #1 \right\| }
\newcommand{\ip}[1]{ \langle #1 \rangle }
\newcommand{\eps}{\varepsilon}
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}
\newcommand{\Pe}{\mathrm{Pe}}

@@

The first problem discussed here is the erosion of a porous medium, as has been preliminarily investigated in \cite{Quaife2018}. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, erosion results from surface shear stress
\begin{align}
\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations \eqref{StokesEq}. 
The motion of the interface is proportional to the absolute value of shear
\begin{align}
\Vn = \CE \, \abs{\tau},
\end{align}
In the latter case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
\label{Ceq}
& \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C \, ,
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \left. \Diff \pd{C}{\ny} \right. %\right\vert_{\bdry} \, ,
\end{equation}
Both processes have been investigated by the PI and collaborators in the context of single bodies in high Reynolds-number flows \cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the low-Reynolds regime of porous-media flow, numerical methods have been developed by the PI and collaborators to simulate shear-induced erosion \cite{Quaife2018}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques. The boundary integral solver is accelerated by the fast multipole method \cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes.
%A near-term goal is to adapt these methods to handle the case of chemically-induced dissolution.

@@

%---------------------------------------------------------------------------------------------------%
\subsection{Transport through eroded media}
\begin{wrapfigure}[]{r}{0.5\textwidth}
  \includegraphics[width=0.5\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel.}
\end{wrapfigure}
Transport through eroded media is a direct consequence of the pore
structure. Qualitative properties of an eroded media, such as channels
of high porosity can be modest in volume fraction, but still be
responsible for a large portion of the flux~\cite{Quaife2018}. To
increase understanding of flow through eroded media, we will simulate
and statistically analyze transport through eroded geometries.

In an eroded geometry $\Omega$, a concentration non-dimensionalized
concentration $c$ is governed by
\begin{align}
  \pd{c}{t} + \uu \cdot \nabla c + \frac{1}{\Pe} \Delta c, 
\end{align}
where $\Pe$ is the dimensionless Peclet number. Here we assume that the
concentration $c$ is affected by the eroded geometry, but the geometry
is not altered further by the concentration. In the infinite Peclet
number limit, the concentration is governed by pure advection and the
characteristics of the concentration field are identical to the
streamlines (Figure~\ref{fig:100tracers}).
\begin{wrapfigure}[]{l}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/tort_eulerian100}
  \caption{\label{fig:100tortuosity} The tortuosity of an eroding
  geometry. The tortuosity can be computed using two different formulas
  (red vs.~blue marks). The red square is the tortuosity in
  Figure~\ref{fig:100tracers}. Their is qualitative agreement between
  our numerical simulations and a power-law model (black dashed
  line)~\cite{mat-kha-koz2008}.}
\end{wrapfigure}

The $j^{th}$ trajectories, or characteristics, is governed by
$\dot{ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed,
using, for example, a high-order Runge-Kutta method, a statistical
analysis be used to describe the transport. Two of the most important
measurements are the tortuosity and the anomalous dispersion, both which
are defined in terms of the length of the trajectories
\begin{align}
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Once $N_p$ trajectories are computed, the tortuosity is the
dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Figure~\ref{fig:100tortuosity}).

While the tortuosity characterizes the amount of winding of streamlines,
the dispersion characterizes the spreading of the trajectories.
Dispersion can be descried using the first- and second-ensemble moments
of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}

%---------------------------------------------------------------------------------------------------%
\subsection{Sinkhole modeling}
